---
title: "Project"
author: "Victor Salit"
date: "Sunday, February 22, 2015"
output: html_document
---


## Executive summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
In this project, our goal is to use data from accelerometers to predict how well the test participants
were doing the exercise.

## Data 
The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

```{r}
#library(knitr)
# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
trainingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the files
if (!file.exists("./data/pml-training.csv")){
    download.file(trainingUrl, destfile = "./data/pml-training.csv")
    }
if (!file.exists("./data/pml-testing.csv")){
    download.file(testingUrl, destfile = "./data/pml-testing.csv")
    }

# reading the data
# thanks go to the forum contributors in the following thread:
# https://class.coursera.org/predmachlearn-011/forum/thread?thread_id=118
train  <- read.csv("./data/pml-training.csv", header=TRUE, na.strings = c("", " ","NA"))
test  <- read.csv("./data/pml-testing.csv", header=TRUE, na.strings = c("", " ","NA"))
```

## Cleaning the data

A brief visual inspection of the column names reveals that the first seven columns are irrelevant for the modelling. All column names in both datasets are equal except the last column. We extract only the columns with direct measurements ('accel', 'magnet' and 'gyros' in their names). Since columns with 'var' in their names contain only 'NA', they are also discarded.

```{r}
namestrain <- colnames(train)
namestest <- colnames(test)
which(namestest!=namestrain)
varind <- grepl('var',namestrain)
accind <- grepl('acc',namestrain)
gyroind <- grepl('gyro',namestrain)
magind <- grepl('magnet',namestrain)
trainMeasured <- cbind(classe = train$classe,train[,(accind|gyroind|magind)&!varind])
testMeasured <- cbind(pid = test$problem_id,test[,(accind|gyroind|magind)&!varind])
names(trainMeasured)
```

## Model fitting
Since the testing dataset includes only 20 observations, we can use the larger dataset to train and crossvalidate the model, whereas the smaller one will be used for the final verification only. We use two methods: decision trees and random forest with default parameters. 

```{r}
# split the cleaned testing data into training and cross validation
library(caret)
indTrain <- createDataPartition(y = trainMeasured$classe, p = 0.6, list = FALSE)
training <- trainMeasured[indTrain, ]
crossval <- trainMeasured[-indTrain, ]

# fit a randomForest model 
library(randomForest)
rfmodel <- randomForest(classe ~ ., data = training) #, preProcess=c("center", "scale")
rfmodel
rfpredictions <- predict(rfmodel, crossval, type = "class")
confusionMatrix(rfpredictions, crossval$classe)

# decision tree model
library(rpart)
dtmodel <- rpart(classe ~ ., data=training, method = "class")
dtmodel
dtpredictions <- predict(dtmodel, crossval, type = "class")
confusionMatrix(dtpredictions, crossval$classe)
```
The random forest model is clearly better (higher accuracy, smaller out of sample error) than the decision tree model. 

Now we can verify it using the original test dataset:

```{r}
verif <- predict(rfmodel, testMeasured, type = "class")
verif

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(verif)
```
All predicted values are correct.