---
title: "Project"
author: "Victor Salit"
date: "Sunday, February 22, 2015"
output: html_document
---


## Executive summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
In this project, our goal is to use data from accelerometers to predict how well the test participants
were doing the exercise.

## Data 
The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

```{r}
# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
trainingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the files
download.file(trainingUrl, destfile = "./data/pml-training.csv")
download.file(testingUrl, destfile = "./data/pml-testing.csv")

# reading the data
# thanks go to the forum contributors in the following thread:
# https://class.coursera.org/predmachlearn-011/forum/thread?thread_id=118
train  <- read.csv("./data/pml-training.csv", header=TRUE, na.strings = c("", " ","NA"))
test  <- read.csv("./data/pml-testing.csv", header=TRUE, na.strings = c("", " ","NA"))
```

## Cleaning and tidying the data

A brief visual inspection of the column names reveals that the first seven columns are irrelevant for the modelling. All column names in both datasets are equal except the last column. We extract only the columns with 'accel' and 'gyros' in their names as primary measurements. 

```{r}
namestrain <- colnames(train)
namestest <- colnames(test)
which(namestest!=namestrain)
varind <- grepl('var',namestrain)
accind <- grepl('acc',namestrain)
gyroind <- grepl('gyro',namestrain)
trainMeasured <- cbind(classe = train$classe,train[,(accind|gyroind)&!varind])
names(trainMeasured)
